peter_models

### 2024-12-08 peter_models.qmd
pushing MLR model update

```{r}
# generate predictions for the ordinal model

new_dataset <- data.frame

predict(ord_mod, new_dataset)
```

@@

```{r}
# generate predictions for the ordinal model with interactions

new_dataset <- data.frame

predict(ord_mod, new_dataset)
```

@@

# Fit multinomial model
m_mult <- multinom(Q29_unordered ~ Q271 + race_recoded + age_recoded +
            # rural + # Rural or Urban/Suburban
            # sex_orientation + # Sexual Orientation
            educ + # Educational Attainment
            S3B + # Gender
            # S7 + # Nativity
            Q21, # Party affiliation
          data = dataset_clean)

@@

# Calculate marginal effects by race
races <- unique(dataset_clean$race_recoded)
marg_effects <- list()

for(race in races) {
    probs <- matrix(0, nrow = length(x_vals), ncol = length(levels(dataset_clean$Q29_unordered)))
    for(i in seq_along(x_vals)) {
        temp_data <- dataset_clean
        temp_data$Q271_simple <- x_vals[i]
        temp_data$race_major <- race
        probs[i,] <- get_avg_probs(m_mult, temp_data, "Q271_simple")
    }
    marg_effects[[race]] <- data.frame(
        x = x_vals,
        race = race,
        prob1 = probs[,1],
        prob2 = probs[,2],
        prob3 = probs[,3],
        prob4 = probs[,4]
    )
}

@@

print("Classification accuracy:")
print(accuracy)
print("Confusion Matrix:")
print(conf_matrix)

# Print AIC
print(AIC(m_mult))

# Calculate predicted probabilities
pred_data <- expand.grid(
    Q271 = unique(dataset_clean$Q271),
    age_recoded = unique(dataset_clean$age_recoded),
    race_recoded = unique(dataset_clean$race_recoded),
    educ = unique(dataset_clean$educ),
    S3B = unique(dataset_clean$S3B),
    Q21 = unique(dataset_clean$Q21)
)

# Get predictions
preds <- predict(m_mult, newdata = pred_data, type = "probs")
pred_data <- cbind(pred_data, as.data.frame(preds))

@@

# Fitting OLR Model Using clm() (Following Jennifer's method)

```{r}
dataset_clean$Q271_simple <- as.numeric(dataset_clean$Q271)
# dataset_clean$Q271 <- factor(dataset_clean$Q271,
#                             levels = c("(5) Not at all important",
#                                      "(4) Slightly important",
#                                      "(3) Moderately important",
#                                      "(2) Very important",
#                                      "(1) Extremely important"),
#                             ordered = TRUE)

m1 <- clm(Q29 ~ Q271 + race_recoded + age_recoded +
            rural + # Rural or Urban/Suburban
            sex_orientation + # Sexual Orientation
            Religion + 
            educ + # Educational Attainment
            S3B + # Gender
            S7 + # Nativity
            Q21, # Party affiliation
          data = dataset_clean)

summary(m1)

# Check for multicollinearity using a linear approximation
vif_m1 <- vif(lm(as.numeric(Q29) ~ Q271 + race_recoded + age_recoded +
            rural + # Rural or Urban/Suburban
            sex_orientation + # Sexual Orientation
            Religion + 
            educ + # Educational Attainment
            S3B + # Gender
            S7 + # Nativity
            Q21, # Party affiliation
          data = dataset_clean))

print("Variance Inflation Factors:")
print(vif_m1)

# Test proportional odds assumption
print("Testing proportional odds assumption:")
test1 <- nominal_test(m1)
print("Nominal test for Model 1:")
print(test1)

# Alternative diagnostic approach using linear approximation
lm_approx <- lm(as.numeric(Q29) ~ Q271 + race_recoded + age_recoded +
            rural + # Rural or Urban/Suburban
            sex_orientation + # Sexual Orientation
            Religion + 
            educ + # Educational Attainment
            S3B + # Gender
            S7 + # Nativity
            Q21, # Party affiliation
          data = dataset_clean)
diagnostics_df <- data.frame(
  fitted = fitted(lm_approx),
  residuals = residuals(lm_approx),
  index = 1:nrow(dataset_clean)
)

# Plot residuals vs fitted values using linear approximation
p1 <- ggplot(diagnostics_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "loess") +
  theme_minimal() +
  labs(title = "Residuals vs Fitted Values (Linear Approximation)",
       x = "Fitted Values",
       y = "Residuals")
print(p1)

# Calculate and plot Cook's distance
cooks_d <- cooks.distance(lm_approx)
influence_df <- data.frame(
  index = 1:length(cooks_d),
  cooks_distance = cooks_d
)

p2 <- ggplot(influence_df, aes(x = index, y = cooks_distance)) +
  geom_point() +
  geom_hline(yintercept = 4/nrow(dataset_clean), color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Cook's Distance Plot",
       x = "Observation Number",
       y = "Cook's Distance")
print(p2)

# Count influential observations
influential_obs <- which(cooks_d > 4/nrow(dataset_clean))
print("Number of influential observations:")
print(length(influential_obs))

# Create prediction visualization
pred_data <- expand.grid(
  Q271 = levels(dataset_clean$Q271),
  race_recoded = levels(dataset_clean$race_recoded), 
  age_recoded = levels(dataset_clean$age_recoded),
  rural = levels(dataset_clean$rural),
  sex_orientation = levels(dataset_clean$sex_orientation),
  Religion = levels(dataset_clean$Religion),
  educ = levels(dataset_clean$educ),
  S3B = levels(dataset_clean$S3B),
  S7 = levels(dataset_clean$S7),
  Q21 = levels(dataset_clean$Q21)
)

# Get predicted probabilities
preds <- predict(m1, newdata = pred_data, type = "prob")
# Convert predictions to numeric scores (weighted average)
pred_data$predicted_score <- apply(preds$fit, 1, function(x) sum(x * 1:4))

# Create prediction plot
p3 <- ggplot(pred_data, aes(x = Q271, y = predicted_score, color = race_recoded)) +
  geom_point() +
  geom_line(aes(group = race_recoded)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Predicted Political Interest by Racial Identity Importance and Race",
       x = "Importance of Racial Identity",
       y = "Predicted Political Interest Level (1-4)",
       color = "Race/Ethnicity")
print(p3)

# Print the coefficients table
print("Coefficients for Model 1:")
coef_table <- summary(m1)$coefficients
print(coef_table)

# Calculate and print odds ratios for the coefficients
odds_ratios <- exp(coef_table[, "Estimate"])
print("Odds Ratios:")
print(odds_ratios)

1.  Model Fit:
    1.  AIC is 42342.54
    2.  Log-likelihood is -21125.27
2.  Proportional Odds Assumption

-   The nominal test for Model 1 shows significant violations (p \< 0.05 for all predictor variables except for Religion)

-   This suggests the effect of racial identity importance varies across different levels of political interest

1.  Multicollinearity

-   Interpretation of VIF values

1.  Residuals and Model Diagnostics

-   The residuals plot shows:

    -   A relatively flat line around zero

    -   Straight downward slope patterns in the residuals

1.  Influential Points:
    -   There are 804 influential observations.

    -   Cook's distance plot shows multiple observations above the threshold.

    -   Given the large sample size (17,545), this number of influential points is not extremely concerning.
2.  Key Findings from Model 1:
    1.  The different levels of racial identity importance are all significant
    2.  There are differences among racial/ethnic groups
3.  Predicted Relationships (from plot)
    1.

>>>>

```{r}
# generate new dataset to create predictions for the models
# 
new_dataset <- data.frame(race_recoded=c("White", "Black", "Asian American", "Latino", "AIAN", "NHPI", "Arab/MENA"),
                          Q271_reordered=c("(5) Not at all important", "(4) Slightly important", "(3) Moderately important", "(2) Very important", "(1) Extremely important", "(4) Slightly important", "(5) Not at all important"),
                          age_recoded = c("18-29", "30-39", "40-49", "50-59", "60-69", "70+", "18-29"),
                          S3B = c("(1) Man", "(2) Woman","(3) Non-binary","(4) Something else (Specify)", "(1) Man", "(2) Woman", "(1) Man"),
                          educ = c("No college experience", "Some college experience", "College graduate", "No college experience", "Some college experience", "College graduate", "College graduate"),
                          Q21 = c("(1) Republican",  "(2) Democrat","(3) Independent", "(4) Other party", "(1) Republican",  "(2) Democrat",  "(3) Independent"))

@@

@@

# Fit multinomial model
m_mult <- multinom(Q29_unordered ~ Q271 * S3B + 
                     race_recoded + 
                     age_recoded +
            educ + # Educational Attainment
            Q21, # Party affiliation
          data = dataset_clean)

@@

# Calculate marginal effects by race
races <- unique(dataset_clean$race_recoded)
marg_effects <- list()
# 
for(race in races) {
    probs <- matrix(0, nrow = length(x_vals), ncol = length(levels(dataset_clean$Q29_unordered)))
    for(i in seq_along(x_vals)) {
        temp_data <- dataset_clean
        temp_data$Q271_simple <- x_vals[i]
        temp_data$race_major <- race
        probs[i,] <- get_avg_probs(m_mult, temp_data, "Q271_simple")
    }
    marg_effects[[race]] <- data.frame(
        x = x_vals,
        race = race,
        prob1 = probs[,1],
        prob2 = probs[,2],
        prob3 = probs[,3],
        prob4 = probs[,4]
    )
}

@@

print("Classification accuracy:")
print(accuracy)
print("Confusion Matrix:")
print(conf_matrix)
# 
# Print AIC
print(AIC(m_mult))

# Calculate predicted probabilities
pred_data <- expand.grid(
    Q271 = unique(dataset_clean$Q271),
    age_recoded = unique(dataset_clean$age_recoded),
    race_recoded = unique(dataset_clean$race_recoded),
    educ = unique(dataset_clean$educ),
    S3B = unique(dataset_clean$S3B),
    Q21 = unique(dataset_clean$Q21)
)
# 
# Get predictions
preds <- predict(m_mult, newdata = pred_data, type = "probs")
pred_data <- cbind(pred_data, as.data.frame(preds))

@@

```{r}
# generate predictions for the ordinal model and the ordinal model  with interactions
predict(ord_mod, new_dataset, type="probs")
predict(ord_mod_interactions, new_dataset, type="probs")

new_dataset <- data.frame(race_recoded=c("White", "Black", "Asian American", "Latino", "AIAN", "NHPI", "Arab/MENA"),
                          Q271=c("(5) Not at all important", "(4) Slightly important", "(3) Moderately important", "(2) Very important", "(1) Extremely important", "(4) Slightly important", "(5) Not at all important"),
                          age_recoded = c("18-29", "30-39", "40-49", "50-59", "60-69", "70+", "18-29"),
                          S3B = c("(1) Man", "(2) Woman","(3) Non-binary","(4) Something else (Specify)", "(1) Man", "(2) Woman", "(1) Man"),
                          educ = c("No college experience", "Some college experience", "College graduate", "No college experience", "Some college experience", "College graduate", "College graduate"),
                          Q21 = c("(1) Republican",  "(2) Democrat","(3) Independent", "(4) Other party", "(1) Republican",  "(2) Democrat",  "(3) Independent"))

predict(m_mult, new_dataset, type = "probs")
```
